### 2025-04-17 Manuscript_Assignment.qmd
commit

---
title: "Manuscript Assignment = Calcasieu"
author: "Nicolas Rivas"
format: 
 html:
    code-fold: true
    code-summary: "Show code"
    toc: true            # Adds a table of contents
    toc-depth: 3         # Sets TOC depth to 3 levels
    number-sections: true # Enables numbered sections
editor: visual
---

```{r Load Libraries}
#| echo: FALSE
#| include: FALSE

library(tidyverse)
library(hypervolume)
library(MuMIn)

```

```{r load data}
#| echo: TRUE
#| include: FALSE

df = read.csv("data/LDWFseinedata.csv") |> 
  filter(basin == "Calcasieu") |> 
  rename_with(tolower) |> 
  mutate(across(where(is.character), tolower),
         year = substr(date, 1,4),
         month = substr(date, 5,6),
         day = substr(date, 7,8)) |>  
  rename(turbidity = tubidity) |> 
  select(-date)

df2 = readRDS('data/first_hvs.rds') |> 
  mutate(year=as.numeric(year))

df_ov=read_csv("data/first_centDist.csv")

dfs_hvs = readRDS('data/first_hvs_species.rds')

suitability_df = readRDS('data/suitability_df.rds')
```

# Create HVs

```{r create HVs}
#| eval: FALSE
#| echo: TRUE

set.seed(1)

## Scaling ####
df2 <- df |> 
  group_by(lat, lon, year, month, day) |> 
  slice(1) |> 
  ungroup() |>
  mutate(across(salinity:airtemp, scale)) |> 
  select(year,salinity, watertemp, turbidity,airtemp) |> 
  group_by(year) |> 
  nest()

head(df2)

## HVs Creation ####

df2 = df2 |> 
  mutate(hv = map(data, \(data) hypervolume_gaussian(data, name = paste(year),
                                                     samples.per.point = 1000,
                                                     kde.bandwidth = estimate_bandwidth(data), 
                                                     sd.count = 3, 
                                                     quantile.requested = 0.95, 
                                                     quantile.requested.type = "probability", 
                                                     chunk.size = 1000, 
                                                     verbose = F)),
         centroid = map(hv, \(hv) get_centroid(hv)),
         size = map_dbl(hv, \(hv) get_volume(hv)))

#saveRDS(df2, 'data/first_hvs.rds')

```

# Create Yearly Comparisons

```{r yearly comparisons}
#| eval: FALSE
#| echo: TRUE

# comparison of across each year
df_y= tibble(y1 = unique(df2$year),
             y2 = unique(df2$year)) |> 
  expand(y1,y2)

# make all unique year comparisons 
df_y = df_y[!duplicated(t(apply(df_y,1,sort))),] %>% 
  filter(!(y1 == y2))

# make two df to join all unique comparisons  
df3 = df2 |> 
  select( y1 = year, hv1 = hv, hv1_size = size, cent1 = centroid)

df4 = df2 |> 
  select( y2 = year, hv2 = hv, hv2_size = size, cent2 = centroid)

# create data frame of all data and make yearly comparisons
df_ov = tibble(year = rep(unique(df2$year),
                           each = nrow(df_y)),
               y1 = rep(df_y$y1, times = length(unique(df2$year))),
               y2 = rep(df_y$y2, times = length(unique(df2$year)))) |> 
  inner_join(df3, by = c('y1')) |> 
  inner_join(df4, by = c('y2')) |> 
  mutate(ychange = y2-y1,
         # calculate the differnces in size 
         lsr = log(hv2_size/hv1_size),
         # join hypervolumees in a set for cverlap
         set = map2(hv1,hv2, \(hv1, hv2) hypervolume_set(hv1, hv2, check.memory = F, verbose = F)),
         # calculate overlap
         ov = map(set, \(set) hypervolume_overlap_statistics(set)),
         # calculate centroid distance 
         dist_cent = 
           map2_dbl(hv1, hv2, \(hv1,hv2) hypervolume_distance(hv1, hv2, type = 'centroid', check.memory=F))) |> 
  #unnest centroid differences
  unnest_wider(ov) |> 
  # select only metrics of interest
  select(year, y1, y2, ychange,lsr,
         dist_cent, jaccard, sorensen,
         uniq_y1 = frac_unique_1, uniq_y2 = frac_unique_2)

beepr::beep(2)

# save output
#write_csv(df_ov, 'data/first_centDist.csv')

```

# Temporal stability

## Overlap

Overlap is calculated using the Sorensen index, which is a measure of similarity between two hypervolumes. The Sorensen index ranges from 0 to 1, where 0 indicates no overlap and 1 indicates complete overlap. We see a linear negative trend across time, indicating that the overlap between hypervolumes decreases.

```{r temporal stability - overlap}
#| echo: TRUE

# overlap 
df_o = df_ov |> 
 # group_by(basin) |>
  nest() |> 
  # fit intercept, linear, and quadratic model
  mutate(m_int = map(data, \(df)lm(sorensen~1, data = df)),
         m_lin = map(data, \(df)lm(sorensen~ychange, data = df)),
         m_quad = map(data, \(df)lm(sorensen~ychange + I(ychange^2), data = df)),
         AICc_int = map_dbl(m_int, \(x) AICc(x)),
         AICc_lin = map_dbl(m_lin, \(x) AICc(x)),
         AICc_quad = map_dbl(m_quad, \(x) AICc(x)),
         model = case_when(
           AICc_int - min(c(AICc_int,AICc_lin,AICc_quad)) <= 4 ~ 'Intercept',
           AICc_lin - AICc_quad <= 4 ~ 'Linear',
           T ~ 'Quadratic'))

# unnest data 
d_o = df_o |> 
  select(data, model) |> 
  unnest(cols = c(data))

ggplot(d_o, aes(ychange, sorensen))+
  geom_point(size = 2.5)+
  geom_smooth(data = d_o |> filter(model == 'Intercept'),
              method = 'lm', formula = y~1, 
              linewidth = 1, color = 'black')+
  geom_smooth(data = d_o |> filter(model == 'Linear'),
              method = 'lm', formula = y~x, 
              linewidth = 1, color = 'blue')+
  geom_smooth(data = d_o |> filter(model == 'Quadratic'),
              method = 'lm', formula = y~x+I(x^2), 
              linewidth = 1, color = 'black')+
  labs(x = 'Years between comparison', y = 'Overlap')+
  scale_color_viridis_d(option = 'turbo')+
  theme_bw()+
  theme(axis.title = element_text(size = 14), 
        axis.text.y = element_text(size = 14, colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"), 
        plot.title = element_text(size = 14, hjust=0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'none',
        legend.title = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.text = element_text(size = 12))

```

## Centroid Distance

Centroid distance increased with time, indicating that the centroid of the hypervolume shifted over time. The linear model was the best fit for the data, indicating a linear increase in centroid distance over time.

```{r temporal stability - Centroid Distance}
#| echo: TRUE

# centroid distance
df_cd = df_ov |> 
  nest() |> 
  # fit intercept, linear, and quadratic model
  mutate(m_int = map(data, \(df)lm(dist_cent~1, data = df)),
         m_lin = map(data, \(df)lm(dist_cent~ychange, data = df)),
         m_quad = map(data, \(df)lm(dist_cent~ychange + I(ychange^2), data = df)),
         AICc_int = map_dbl(m_int, \(x) AICc(x)),
         AICc_lin = map_dbl(m_lin, \(x) AICc(x)),
         AICc_quad = map_dbl(m_quad, \(x) AICc(x)),
         model = case_when(
           AICc_int - min(c(AICc_int,AICc_lin,AICc_quad)) <= 4 ~ 'Intercept',
           AICc_lin - AICc_quad <= 4 ~ 'Linear',
           T ~ 'Quadratic'))

# unnest data 
d_cd = df_cd |> 
  select(data, model) |> 
  unnest(cols = c(data))

ggplot(d_cd, aes(ychange, dist_cent))+
  geom_hline(aes(yintercept = 1), linetype = 'dashed')+
  geom_point(size = 2.5)+
  geom_smooth(data = d_cd |> filter(model == 'Intercept'),
              method = 'lm', formula = y~1, 
              linewidth = 1, color = 'blue')+
  geom_smooth(data = d_cd |> filter(model == 'Linear'),
              method = 'lm', formula = y~x, 
              linewidth = 1, color = 'blue')+
  geom_smooth(data = d_cd |> filter(model == 'Quadratic'),
              method = 'lm', formula = y~x+I(x^2), 
              linewidth = 1, color = 'blue')+
  labs(x = 'Years between comparison', y = 'Centroid distance')+
  scale_color_viridis_d(option = 'turbo')+
  theme_bw()+
  theme(axis.title = element_text(size = 14), 
        axis.text.y = element_text(size = 14, colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"), 
        plot.title = element_text(size = 14, hjust=0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'none',
        legend.title = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.text = element_text(size = 12))

```

## Log Size Ratio

The log size ratio (LSR) is a measure of the change in size of the hypervolume over time. A LSR of 1 indicates no change in size, while a LSR greater than 1 indicates an increase in size and a LSR less than 1 indicates a decrease in size. The LSR decreased over time, indicating that the hypervolume shrank over time. The quadratic model was the best fit for the data, indicating a non-linear decrease in size over time.

```{r temporal stability - log size ratio}
#| echo: TRUE

# log size ratio
df_lsr = df_ov |> 
  nest() |> 
  # fit intercept, linear, and quadratic model
  mutate(m_int = map(data, \(df)lm(lsr~1, data = df)),
         m_lin = map(data, \(df)lm(lsr~ychange, data = df)),
         m_quad = map(data, \(df)lm(lsr~ychange + I(ychange^2), data = df)),
         AICc_int = map_dbl(m_int, \(x) AICc(x)),
         AICc_lin = map_dbl(m_lin, \(x) AICc(x)),
         AICc_quad = map_dbl(m_quad, \(x) AICc(x)),
         model = case_when(
           AICc_int - min(c(AICc_int,AICc_lin,AICc_quad)) <= 4 ~ 'Intercept',
           AICc_lin - AICc_quad <= 4 ~ 'Linear',
           T ~ 'Quadratic'))

# unnest data 
d_lsr = df_lsr |> 
  select(data, model) |> 
  unnest(cols = c(data))

ggplot(d_lsr, aes(ychange, lsr))+
  geom_hline(aes(yintercept = 1), linetype = 'dashed')+
  geom_point(size = 2.5)+
  geom_smooth(data = d_lsr |> filter(model == 'Intercept'),
              method = 'lm', formula = y~1, 
              linewidth = 1, color = 'blue')+
  geom_smooth(data = d_lsr |> filter(model == 'Linear'),
              method = 'lm', formula = y~x, 
              linewidth = 1, color = 'blue')+
  geom_smooth(data = d_lsr |> filter(model == 'Quadratic'),
              method = 'lm', formula = y~x+I(x^2), 
              linewidth = 1, color = 'red')+
  labs(x = 'Years between comparison', y = 'log(y2 size/y1 size)')+
  scale_color_viridis_d(option = 'turbo')+
  theme_bw()+
  theme(axis.title = element_text(size = 14), 
        axis.text.y = element_text(size = 14, colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"), 
        plot.title = element_text(size = 14, hjust=0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'none',
        legend.title = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.text = element_text(size = 12))

```

# Species

Using white shrimp, blue crab, brown shrimp

1.  White Shrimp (Litopenaeus setiferus) Estuarine-dependent species Spawns offshore, juveniles migrate into estuaries Highly sensitive to salinity and temperature

2.  Brown Shrimp (Farfantepenaeus aztecus) — Closely related competitor Similar habitat use to white shrimp Slightly different seasonal and salinity preferences Useful for testing niche differentiation or competitive overlap

3.  Blue Crab (Callinectes sapidus) — Functionally different predator Also estuarine-dependent Omnivorous predator, unlike shrimp Occupies broader salinity/temperature range Useful for testing niche separation between trophic levels

## First filter, scale and create HVs

```{r species}
#| echo: TRUE
#| eval: FALSE

dfs_hvs <- df |>
  filter(species %in% c("white shrimp", "blue crab", "brown shrimp")) |>
  mutate(across(c(salinity, watertemp, turbidity, airtemp), scale)) |>
  select(species, salinity, watertemp, turbidity, airtemp, num) |>
  group_by(species) |>
  nest(weight = num, data = salinity:airtemp) |> 
  mutate(hv = map2(data,weight, \(data,weight) hypervolume_gaussian(data, name = paste(species),
                                                     samples.per.point = 1000,
                                                     weight = weight$num,
                                                     kde.bandwidth = estimate_bandwidth(data), 
                                                     sd.count = 3, 
                                                     quantile.requested = 0.95, 
                                                     quantile.requested.type = "probability", 
                                                     chunk.size = 1000, 
                                                     verbose = F)),
         centroid = map(hv, \(hv) get_centroid(hv)),
         size = map_dbl(hv, \(hv) get_volume(hv)))

beepr::beep(3)

```

```{r species - plot hvs}
#| echo: TRUE

hvj = hypervolume_join(dfs_hvs$hv[[1]], dfs_hvs$hv[[2]], dfs_hvs$hv[[3]])
plot(hvj,
     colors = c('#0000CC', '#D72638', '#228B22'))

```

```{r species - comparisons}
#| echo: TRUE
#| eval: FALSE

# comparison of across each species
df_y= tibble(y1 = unique(dfs_hvs$species),
             y2 = unique(dfs_hvs$species)) |> 
  expand(y1,y2)

# make all unique species comparisons 
df_y = df_y[!duplicated(t(apply(df_y,1,sort))),] %>% 
  filter(!(y1 == y2))

# make two df to join all unique comparisons  
df1 = dfs_hvs |> 
  select(y1 = species, hv1 = hv, hv1_size = size, cent1 = centroid)

df2 = dfs_hvs |> 
  select(y2 = species, hv2 = hv, hv2_size = size, cent2 = centroid)

# create data frame of all data and make species comparisons
df_ov = tibble(species = rep(unique(dfs_hvs$species),
                           each = nrow(df_y)),
               y1 = rep(df_y$y1, times = length(unique(dfs_hvs$species))),
               y2 = rep(df_y$y2, times = length(unique(dfs_hvs$species)))) |> 
  inner_join(df1, by = c( 'y1')) |> 
  inner_join(df2, by = c( 'y2')) |> 
  mutate(
         # calculate the differnces in size 
         lsr = log(hv2_size/hv1_size),
         # join hypervolumees in a set for cverlap
         set = map2(hv1,hv2, \(hv1, hv2) hypervolume_set(hv1, hv2, check.memory = F, verbose = F)),
         # calculate overlap
         ov = map(set, \(set) hypervolume_overlap_statistics(set)),
         # calculate centroid distance 
         dist_cent = map2_dbl(hv1, hv2, \(hv1,hv2) hypervolume_distance(hv1, hv2, type = 'centroid', check.memory=F))) |> 
  #unnest centroid differences
  unnest_wider(ov) |> 
  # select only metrics of interest
  select(y1, y2,lsr,
         dist_cent, jaccard, sorensen,
         uniq_y1 = frac_unique_1, uniq_y2 = frac_unique_2)

# save output
#write_csv(df_ov, 'data/species_centDist.csv')

```

# Suitability

## Load Data make comparisons

```{r suitability}
#| echo: TRUE
#| eval: FALSE
hv_species = dfs_hvs |> 
  select(species, hv)

# Keep only environmental data
env_scaled = df |>
  filter(species %in% c("white shrimp", "blue crab", "brown shrimp")) |>
  select(year, salinity, watertemp, turbidity, airtemp) |>
  drop_na() |>
  mutate(across(everything(), as.numeric)) |>
  mutate(across(salinity:airtemp, scale)) |> 
  group_by(year) |> 
  nest()

# Expand grid of species × years
suitability_df <- crossing(
  species = c("white shrimp", "blue crab", "brown shrimp"),
  year = unique(env_scaled$year)
)

# Join HVs and data
suitability_df <- suitability_df |>
  left_join(hv_species, by = "species") |>
  left_join(env_scaled, by = "year")

# Calculate % of suitable environments
suitability_df <- suitability_df |>
  mutate(prop_suitable = map2_dbl(hv, data, ~{
    points_inside <- hypervolume::hypervolume_inclusion_test(.x, .y)
    mean(points_inside)  # proportion of sites inside the niche
  }))

# Caused by warning in `hypervolume::hypervolume_inclusion_test()`:
#   ! Results may have a high error rate. Consider setting fast.or.accurate='accurate'.
```

Habitat suitability seems to remain stable for blue crabs, fluctuates for brown shrimp but generally stable and decreases for white shrimp near 2018

```{r suitability - plot}
ggplot(suitability_df, aes(x = year, y = prop_suitable, color = species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  labs(x = "Year", y = "Proportion of Suitable Sites",
       title = "Habitat Suitability Over Time")

```

### 2025-04-17 workshop3.R

@@ -58,7 +58,7 @@ head(df)
 # generate hypervolumes
 df = df |> 
   mutate(hv = map(data, \(data) hypervolume_gaussian(data, name = paste(BASIN,YEAR,sep = '_'),
-                                                     samples.per.point = 1000,
+                                                     samples.per.point = 100,
                                                      kde.bandwidth = estimate_bandwidth(data), 
                                                      sd.count = 3, 
                                                      quantile.requested = 0.95, 

### 2025-04-17 Report.R
c

@@ -406,32 +406,32 @@ ggplot(d, aes(ychange, lsr))+
 
 # 1. White Shrimp (Litopenaeus setiferus)
 # Estuarine-dependent species
-# 
+#
 # Spawns offshore, juveniles migrate into estuaries
-# 
+#
 # Highly sensitive to salinity and temperature
-# 
+#
 # 🔶 2. Brown Shrimp (Farfantepenaeus aztecus) — Closely related competitor
 # Similar habitat use to white shrimp
-# 
+#
 # Slightly different seasonal and salinity preferences
-# 
+#
 # Useful for testing niche differentiation or competitive overlap
-# 
+#
 # 🔷 3. Blue Crab (Callinectes sapidus) — Functionally different predator
 # Also estuarine-dependent
-# 
+#
 # Omnivorous predator, unlike shrimp
-# 
+#
 # Occupies broader salinity/temperature range
-# 
+#
 # Useful for testing niche separation between trophic levels
-# 
+#
 # 🧠 This trio lets you ask:
 #   "To what extent do co-occurring estuarine species with differing ecologies (two shrimp and one predator) share environmental niche space in Calcasieu estuary?"
-# 
+#
 # Or alternatively:
-#   
+#
 #   "Do closely related penaeid shrimp partition environmental niche space, and how does that compare to a more distantly related estuarine predator?"
 
 
@@ -474,12 +474,12 @@ hvj = hypervolume_join(dfs_hvs$hv[[1]], dfs_hvs$hv[[2]], dfs_hvs$hv[[3]])
 plot(hvj)
 
 
-# comparison of across each year
+# comparison of across each species
 df_y= tibble(y1 = unique(dfs_hvs$species),
              y2 = unique(dfs_hvs$species)) |> 
   expand(y1,y2)
 
-# make all unique year comparisons 
+# make all unique species comparisons 
 df_y = df_y[!duplicated(t(apply(df_y,1,sort))),] %>% 
   filter(!(y1 == y2))
 
@@ -491,7 +491,7 @@ df2 = dfs_hvs |>
   select(y2 = species, hv2 = hv, hv2_size = size, cent2 = centroid)
 
 
-# create data frame of all data and make yearly comparisons
+# create data frame of all data and make species comparisons
 df_ov = tibble(species = rep(unique(dfs_hvs$species),
                            each = nrow(df_y)),
[…]

### 2025-04-18 Manuscript_Assignment.qmd
c

@@
 library(MuMIn)
+library(ggpubr)
+library(gt)
 
@@
 
+## HV plots
+
 ```{r species - plot hvs}
@@
 
+## Niche SIze
+
+Nice size of the three species
+
+```{r species - plot size}
+#| echo: FALSE
+
+dfs_hvs |>
+  select(species, size) |>
+  ggplot(aes(x = species, y = 1, size = size, fill = species)) +
+  geom_point(shape = 21, color = "black") +
+  geom_text(aes(label = round(size, 2)), vjust = -3, size = 9) +  # nicely formatted labels
+  scale_size_continuous(range = c(10, 40)) +
+  labs(
+    title = "Niche Size of Estuarine Species",
+    x = "Species",
+    y = NULL
+  ) +
+  guides(size = "none", fill = "none") +  # remove legends
+  theme_classic(base_size = 14) +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major.y = element_blank()
+  )
+```
+
+## Overlap
+
+```{r species - table overlap}
+#| echo: FALSE
+#| eval: FALSE
+
+df_ov_dedup <- df_ov |> 
+  group_by(y1, y2) |> 
+  summarise(across(where(is.numeric), mean), .groups = "drop")
+
+df_ov_dedup |>
+  mutate(across(where(is.numeric), ~ round(., 3))) |>
+  gt() |>
+  tab_header(
+    title = "Pairwise Niche Overlap and Dissimilarity Metrics"
+  ) |>
+  cols_label(
+    y1 = "Species 1",
+    y2 = "Species 2",
+    lsr = "Log Size Ratio",
+    dist_cent = "Centroid Distance",
+    jaccard = "Jaccard",
+    sorensen = "Sorensen",
+    uniq_y1 = "Unique to Species 1",
+    uniq_y2 = "Unique to Species 2"
+  )
+
+```
+![Pairwise Niche Overlap Table](images/niche_overlap_table.png)
+
+Heatmap of similarity between species using Sorensen
+
+![Niche Overlap Heatmap](sorensen_heatmap.png)
+
 # Suitability
@@
 
+## Habitat Suitability/Time
+
 Habitat suitability seems to remain stable for blue crabs, fluctuates for brown shrimp but generally stable and decreases for white shrimp near 2018
@@
 ```
+
+## Species Abundance
+
+I went ahead and plotted abundance data. No brown shrimp or white shimp were caught in 2019 and in general there was a continous decrease in their abundance numbers.
+
+We can see that sharp fluctuation in abundance is normal but did not reach zero before 2019. Panel B just has a closer look at the last three years
+
+```{r - abundance}
+#| echo: TRUE
+
+target_years <- as.character(1986:2019)
+target_species <- c("white shrimp", "brown shrimp", "blue crab")
+
+# Fill in missing combinations with 0
+species_counts <- df |>
+  filter(species %in% target_species,
+         year %in% target_years) |>
+  group_by(year, species) |>
+  summarise(n_caught = sum(num), .groups = "drop") |>
+  complete(year = target_years, species = target_species, fill = list(n_caught = 0)) |>
+  arrange(year, species)
+
+a = ggplot(species_counts, aes(x = as.numeric(year), y = n_caught, color = species)) +
+  geom_line(size = 1.2) +
+  geom_point(size = 2.5) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(
+    title = "Abundance of Estuarine Species (1986–2019)",
+    x = "Year",
+    y = "Number Caught",
+    color = "Species"
+  ) +
+  theme_classic(base_size = 14)
+
+species_counts_zoom <- species_counts |>
+  filter(as.numeric(year) >= 2017 & as.numeric(year) <= 2019)
+
+b = ggplot(species_counts_zoom, aes(x = as.numeric(year), y = n_caught, color = species)) +
+  geom_line(linewidth = 1.2) +
+  geom_point(size = 2.5) +
+  scale_x_continuous(breaks = 2017:2019, labels = 2017:2019) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(
+    title = "Abundance of Estuarine Species (2017–2019)",
+    x = "Year",
+    y = "Number Caught",
+    color = "Species"
+  ) +
+  theme_classic(base_size = 14)
+
+ggarrange(a, b,
+          ncol = 2, nrow = 1,
+          labels = c("A", "B"),
+          common.legend = TRUE,
+          legend = "bottom") +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+## 2019 Suitability
+
+Upon Further investigation.
+we see that the habitat suitbability in 2019 changed outsize of the centroid of white shrimp and brown shrimp.
+It was not longer suitable and had greater changes in magnitude compared to blue crab.....(i think)
+
+```{r - suitability - 2019}
+#| echo: FALSE
+#| message: false
+
+suitability_2019_summary = read_csv("data/suitability_2019_summary.csv")
+
+suitability_2019_summary |> 
+  mutate(across(where(is.numeric), ~ round(., 2))) |>
+  gt() |>
+  tab_header(
+    title = "Habitat Suitability and Niche Deviation (2019)"
+  ) |>
+  cols_label(
+    species = "Species",
+    prop_suitable = "Prop. Suitable",
+    salinity = "Δ Salinity",
+    watertemp = "Δ Water Temp",
+    turbidity = "Δ Turbidity",
+    airtemp = "Δ Air Temp"
+  ) |>
+  fmt_number(everything(), decimals = 2)
+
+```

### 2025-04-18 workshop3.R

@@ -638,3 +638,53 @@ ggplot(df_all, aes(axis, imp_lsr, fill = BASIN))+
         legend.text = element_text(size = 12))
 
 
+plot(dfs_hvs$hv[[1]])
+plot(dfs_hvs$hv[[2]])
+plot(dfs_hvs$hv[[3]])
+
+
+hvj = hypervolume_join(dfs_hvs$hv[[1]], dfs_hvs$hv[[2]], dfs_hvs$hv[[3]])
+plot(hvj)
+
+# comparison of across each year
+df_y= tibble(y1 = unique(df$YEAR),
+             y2 = unique(df$YEAR)) |> 
+  expand(y1,y2)
+
+# make all unique year comparisons 
+df_y = df_y[!duplicated(t(apply(df_y,1,sort))),] %>% 
+  filter(!(y1 == y2))
+
+# make two df to join all unique comparisons  
+df1 = df |> 
+  select(BASIN, y1 = YEAR, hv1 = hv, hv1_size = size, cent1 = centroid)
+
+df2 = df |> 
+  select(BASIN, y2 = YEAR, hv2 = hv, hv2_size = size, cent2 = centroid)
+
+
+# create data frame of all data and make yearly comparisons
+df_ov = tibble(BASIN = rep(unique(df$BASIN),
+                           each = nrow(df_y)),
+               y1 = rep(df_y$y1, times = length(unique(df$BASIN))),
+               y2 = rep(df_y$y2, times = length(unique(df$BASIN)))) |> 
+  inner_join(df1, by = c('BASIN', 'y1')) |> 
+  inner_join(df2, by = c('BASIN', 'y2')) |> 
+  mutate(ychange = y2-y1,
+  # calculate the differnces in size 
+         lsr = log(hv2_size/hv1_size),
+  # join hypervolumees in a set for cverlap
+         set = map2(hv1,hv2, \(hv1, hv2) hypervolume_set(hv1, hv2, check.memory = F, verbose = F)),
+  # calculate overlap
+         ov = map(set, \(set) hypervolume_overlap_statistics(set)),
+  # calculate centroid distance 
+         dist_cent = map2_dbl(hv1, hv2, \(hv1,hv2) hypervolume_distance(hv1, hv2, type = 'centroid', check.memory=F))) |> 
+  #unnest centroid differences
+  unnest_wider(ov) |> 
+  # select only metrics of interest
[…]

### 2025-04-18 Manuscript_Assignment.qmd
c

@@
 #write_csv(df_ov, 'data/first_centDist.csv')
+
+```
+
+Overlap
+```{r yearly comparisons - plot}
+#| echo: FALSE
+d = df_ov |> 
+  filter(ychange == 1)
+
+# overlap
+ggplot(d, aes(y2, sorensen,color="blue"))+
+  geom_point(size = 2)+
+  geom_line(linewidth = 1)+
+  labs(x = 'Year', y = 'Overlap')+
+  scale_fill_viridis_d(option = 'turbo')+
+  scale_color_viridis_d(option = 'turbo')+
+  theme_bw()+
+  #facet_wrap(~BASIN, nrow = 2)+
+  theme(axis.title = element_text(size = 14), 
+        axis.text = element_text(size = 14, colour = "gray0"), 
+        plot.title = element_text(size = 14, hjust=0.5),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        legend.position = 'none',
+        legend.title = element_text(size = 14),
+        strip.text.x = element_text(size = 14),
+        legend.text = element_text(size = 12))
+
+```
+
+Centroid Distance
+```{r yearly comparisons - plot2}
+#| echo: FALSE
+
+# centroid distance 
+ggplot(d, aes(y2, dist_cent))+
+  geom_hline(aes(yintercept = 1), linetype = 'dashed', linewidth = 1)+
+  geom_point(size = 2)+
+  geom_line(linewidth = 1)+
+  labs(x = 'Year', y = 'Centroid distance')+
+  scale_fill_viridis_d(option = 'turbo')+
+  scale_color_viridis_d(option = 'turbo')+
+  theme_bw()+
+  #facet_wrap(~BASIN, nrow = 2)+
+  theme(axis.title = element_text(size = 14), 
+        axis.text = element_text(size = 14, colour = "gray0"), 
+        plot.title = element_text(size = 14, hjust=0.5),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        legend.position = 'none',
+        legend.title = element_text(size = 14),
+        strip.text.x = element_text(size = 14),
+        legend.text = element_text(size = 12))
+
+```
+
+LSR
+
+```{r yearly comparisons - plot3}
+#| echo: FALSE
+
+# log size ratio
+ggplot(d, aes(y2, lsr))+
+  geom_hline(aes(yintercept = 0), linetype = 'dashed', linewidth = 1)+
+  geom_point(size = 2)+
+  geom_line(linewidth = 1)+
+  labs(x = 'Year', y = 'log(y2 size/y1 size)')+
+  scale_fill_viridis_d(option = 'turbo')+
+  scale_color_viridis_d(option = 'turbo')+
+  theme_bw()+
+  #facet_wrap(~BASIN, nrow = 2)+
+  theme(axis.title = element_text(size = 14), 
+        axis.text = element_text(size = 14, colour = "gray0"), 
+        plot.title = element_text(size = 14, hjust=0.5),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        legend.position = 'none',
+        legend.title = element_text(size = 14),
+        strip.text.x = element_text(size = 14),
+        legend.text = element_text(size = 12))
 
@@
 ```{r - abundance}
-#| echo: TRUE
+#| echo: FALSE
 

### 2025-04-18 Report.R

@@ -462,6 +462,8 @@ beepr::beep(3)
 
 #saveRDS(dfs_hvs, 'data/first_hvs_species.rds')
 
+dfs_hvs=readRDS('data/first_hvs_species.rds')
+
 head(dfs_hvs)
 
 
@@ -520,6 +522,64 @@ df_ov = tibble(species = rep(unique(dfs_hvs$species),
 head(df_ov)
 # Species comparisons
 
+df_ov=read_csv('data/species_centDist.csv')
+
+
+
+
+
+
+dfs_hvs |>
+  select(species, size) |>
+  ggplot(aes(x = species, y = size, fill = species)) +
+  geom_col(width = 0.6, show.legend = FALSE) +
+  labs(
+    title = "Overall Niche Size of Estuarine Species",
+    x = "Species",
+    y = "Hypervolume Size"
+  ) +
+  theme_minimal(base_size = 14)+
+  geom_text(aes(label = round(size, 3)), vjust = -0.3, size = 5)
+
+
+dfs_hvs |>
+  select(species, size) |>
+  ggplot(aes(x = species, y = 1, size = size, fill = species)) +
+  geom_point(shape = 21, color = "black") +
+  geom_text(aes(label = round(size, 2)), vjust = -3, size = 9) +  # nicely formatted labels
+  scale_size_continuous(range = c(10, 40)) +
+  labs(
+    title = "Niche Size of Estuarine Species",
+    x = "Species",
+    y = NULL
+  ) +
+  guides(size = "none", fill = "none") +  # remove legends
+  theme_classic(base_size = 14) +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major.y = element_blank()
+  )
+
+
+df_ov_dedup <- df_ov |> 
+  group_by(y1, y2) |> 
+  summarise(across(where(is.numeric), mean), .groups = "drop")
+
+ggplot(df_ov_dedup, aes(x = y1, y = y2, fill = sorensen)) +
+  geom_tile(color = "white") +
+  geom_text(aes(label = round(sorensen, 2)), size = 4) +
+  scale_fill_viridis_c(option = "magma", direction = -1, limits = c(0, 1)) +
+  labs(
+    title = "Pairwise Niche Overlap (Sorensen Similarity)",
+    x = "Species 1",
+    y = "Species 2",
+    fill = "Sorensen"
+  ) +
[…]

### 2025-04-18 Manuscript_Assignment.qmd
c

@@
 
-Overlap
+## Overlap
+
 ```{r yearly comparisons - plot}
@@
 
-Centroid Distance
+## Centroid Distance
+
 ```{r yearly comparisons - plot2}
@@
 
-LSR
+## LSR
 
@@
 ```
+
 ![Pairwise Niche Overlap Table](images/niche_overlap_table.png)
